### 2015-04-28 thesis-proposal/problem.tex
Updated poropsal

%Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives. %New primitives, and thus new dialects of these languages, should arise very rarely. 
%New primitives (and new \emph{dialects} of the language). %Extending a general-pandurpose language with new primitives, i.e.  forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare. %Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality. %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
%Many languages claim to achieve this ideal. 
%For example, 
%If we measure the generality of a language by how frequently  dialects of the language emerge, 
%all major contemporary languages seem to leave room for improvement. 
%General-purpose programming languages are a dime a. 
%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it. Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in. 
Functional programming languages like Standard ML (SML) and Haskell descend directly from typed lambda calculi, diverging mainly in that they consider  various usability criteria in addition to fundamental metatheoretic issues in their design.  
For example, they build in record types, generalizing the nullary and binary product types more common in simpler calculi, because labeled components are cognitively useful to programmers. Similarly, they both build in derived syntax for the list datatype.% concision and readability. 
%Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML. to be useful in practice, it must present a decidedly more elaborate concrete syntax and type structure to programmers, designed considering various human factors in addition to fundamental metatheoretic issues. For example, while it is quite common to study minimal calculi that build in only nullary and binary product types, Standard ML (SML) builds in record types, because labels are cognitively useful. %syntactic and semantic primitives that capture common idioms naturally and precisely.

Ideally, this transition from a minimal calculus to a pragmatic ``full-scale'' programming language would require only a limited number of primitive ``embellishments'' like these. The community around the language would then be able to direct its efforts exclusively toward the development of useful libraries. Unfortunately, a stable language design that achieves this ideal has yet to emerge, as evidenced by the diverse array of {dialects} that continue to proliferate around all major contemporary languages, functional languages like ML included. % For example, let us consider the many dialects of ML. Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with recursive sum and product types, they all introduce various primitive syntactic and semantic extensions and tweaks of their own design (we cover several examples below). 
%Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
Indeed, tools that assist in the construction of ``domain-specific'' language dialects  seem only to be increasing in popularity. 
{This calls for an investigation}: why are well-informed programmers and researchers so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' functional primitives already available in major languages today?
%Let us investigate why this is.
%What kinds of constructs are not expressible in languages like SML?

@@

Perhaps the most common  reason may be that the \emph{syntactic cost} of expressing a construct of interest in terms of contemporary general-purpose primitives is not always ideal. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}. This  will allow  us to express  precisely the semantics that we seek, but syntactically,  the usual approach of using run-time string parsing whenever we introduce a pattern will leave something to be desired. To avoid run-time string parsing, we will be tempted to introduce derived syntactic forms specific to our library, thereby constructing a \emph{derived syntactic dialect}, i.e. a dialect specified by context-free elaboration to the existing language. Such derived syntactic dialects are not uncommon. For example Ur/Web is a  derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}. %In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g. derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}. 
Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} are designed to lower the engineering costs of constructing derived syntactic dialects, contributing to their continued proliferation (we will consider them further in Sec. \ref{sec:syntax}). % contributing to their proliferation.

@@

More advanced dialects define new static type structure, going beyond what is possible by context-free elaboration to the existing language. Record types themselves are perhaps  the simplest example -- they cannot be expressed by context-free elaboration to a language with only nullary and binary products. Various dialects of ML and Haskell have explored  ``record-like'' primitives that go further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
, methods, prototypic dispatch and various other semantic ``embellishments'' that cannot be expressed by context-free elaboration to a language with standard record types (we will detail an example in Sec. \ref{sec:metamodules}). The Ocaml dialect of ML builds in ``polymorphic variants'', ``open datatypes'' and  logic for typechecking operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}. ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.  ML5 builds in primitives for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}. Manticore builds in parallel arrays \cite{conf/popl/FluetRRSX07}. Alice ML builds in futures and promises \cite{AliceLookingGlass}. MLj builds in the entire Java programming language, to support typesafe interoperation with Java libraries \cite{Benton:1999:IWW:317636.317791}. Tools like proof assistants and logical frameworks support specifying and reasoning metatheoretically about dialects like these, and tools like compiler generators and language workbenches lower the cost of implementing them, contributing to their  proliferation.

@@

% express record types as syntactic sugar over the simply-typed lambda calculus with  binary product types.\footnote{Pairs can of course be expressed as syntactic sugar atop records, though one could argue that using binary products as the more primitive concept is simpler.} The static semantics need to be extended with new type and term operators. However, the simplest way to express the dynamic semantics of the newly introduced term operators is by translation to nested binary products, so we can leave the operational semantics alone. \todo{fill this out} %For example, there are dozens of constructs that go by the name of ``records'' in various languages, each defined by a slightly different collection of primitive operations. \todo{examples} %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}. Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally.

The reason why this proliferation of language dialects should be considered alarming is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect. %At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,  
To do so would fundamentally require combining the two dialects into a single language. Left unconstrained, this is an ill-posed problem, i.e. given two dialects that are specified by judgements of arbitrary form, it is not clear what it would even mean to combine them. Even when the dialects are specified using some formalism that does define what it means to combine two dialects, there is generally no guarantee that the composition will conserve important syntactic and semantic properties. %There is no well-defined mechanism for constructing such a ``combined language'' in general. 
For example, consider two derived syntactic dialects, one building in derived syntax for JSON (a popular data interchange format), the other using a similar syntax for finite mappings of some other type. Though each is known to have an unambiguous concrete syntax in isolation, when their grammars are na\"ively  combined (e.g. by Camlp4), ambiguities arise.  Any putative ``combined language'' must formally be considered a  distinct system for which one must derive all metatheorems of interest anew, guided only informally by those derived for the dialects individually. % would generally attempt to combine their semantics, particularly when they are specified in very different ways. %the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects. 
%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. %A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable. 
%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
Due to this paucity of modular reasoning principles, language dialects are not practical artifacts for software development ``in the large''. %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.

Dialects do, however, have a more indirect impact on large-scale software development in that the language designers that control comparatively popular languages, like Ocaml and Scala, can be convinced to incorporate  ideas from dialects  into backwards compatible language revisions. %These decisions are increasingly influenced by community processes, e.g. the Scala Improvement Process.  %This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g. the Scala Improvement Process). 
%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
However, this has come at a cost: these languages have ballooned in size. This is not a scalable approach -- as suggested by the diversity of examples above, there is a ``long tail'' of potentially useful such language extensions (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}). %rimitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still quite often left languishing in impractical ``toy'' dialects. 
Recalling the words of  Reynolds, which are as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition. %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.

\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
and generality in language design is a serious unsolved 
problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
\end{quote}

\newpage %A measured approach to incorporating new primitives into a general-purpose programming language is  sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below. 
This leaves the language design community with two possible paths forward if we wish to keep general-purpose languages small and free of \emph{ad hoc} primitives. One, exemplified (arguably) by SML, is to generally eschew the introduction of new primitives and settle on a set of primitives that sit at a ``sweet spot'' in the overall language design space, accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost. % that this implies.  
The other path forward is to continue on in the search for even more general language primitives, i.e. primitives that are so expressive that they permit us to degrade a broad class of existing primitives, including those found in dialects, to modularly composable library constructs, so that they can be evaluated on their individual merits by programmers and used together without the possibility of conflict. % This should render the construction of new dialects increasingly unnecessary. 
Encouragingly, primitives that have such a ``slimming'' effect on a language do occasionally arise. For example, a recent revision of Ocaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}. Using GADTs, Ocaml was able to move some of the machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into the standard library (but some machinery remains built in). Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes  more dramatic steps down this second path by allowing library providers to express new derived syntax and type structure of their own design, in a safe and modularly composable manner.

@@

%\vspace{-10px}
\section{Proposed Contributions}

Verse features a module system taken directly from SML, but its core language is organized in a novel manner as a \emph{type-directed translation semantics} targeting a minimal typed lambda calculus called the Verse \emph{internal language} (IL). We formally introduce Verse's core semantics in Sec. \ref{sec:verse}. In lieu of the typical \emph{external} (i.e. programmer-facing) derived syntax and  type structure, Verse introduces two novel primitive constructs:
\begin{itemize}
\item \textbf{Typed syntax macros} (TSMs), introduced in Sec. \ref{sec:syntax}, subsume the need to primitively build in derived concrete syntax specific to library constructs, e.g. list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax. %We show how TSMs can be invoked explicitly by clients, or, implicitly based on local type inference.
\item \textbf{Metamodules}, introduced in Sec. \ref{sec:metamodules}, subsume the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs by giving library providers static hooks directly into the type-directed translation semantics. For example, a metamodule that implements the type structure of records would:
\begin{enumerate}
\item Introduce a type constructor \lstinline{record} parameterized by finite mappings from labels to types, defining its semantics by translation to (for example) unary and binary products (which are built in to the internal language); and 
\item Introduce operators used to work with records, minimally record introduction and elimination, but perhaps also various functional update operators, defining the logic governing their typechecking and translation to the IL directly. 
\end{enumerate}
We will see direct analogies between modules and metamodules as we continue.
\end{itemize} 
Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because their static semantics involves the invocation of user-defined functions.% static code generation. %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.

The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to decentralize control over aspects of the language's concrete syntax and type structure that, in other contemporary languages, are under the exclusive control of the language designer. If we are not careful, many of the problems discussed earlier as inherent to combining distinct language dialects could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer. The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended. We will avoid this somewhat confusing terminology.} Our main technical contributions will come in showing how to address these problems in a principled manner. In particular, syntactic conflicts will be made impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}. In other words, library providers will have the ability to reason about the constructs that they define in isolation, and clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We  assume that naming conflicts can be avoided extrinsically, e.g. by using a URI-based naming scheme as in the Java ecosystem.}

@@

Finally, we are not interested here in languages that feature full-spectrum dependent types, which blur the phase separation between compile-time and run-time, though we conjecture that the primitives we introduce could be introduced into languages like Gallina (the ``external language'' of the Coq proof assistant) with some modifications. Verse should be compared to languages that maintain a phase separation, like ML, Haskell and Scala. %Our interest is in making sure that the standard library is no more privileged than any other library. % Our interest is not in creating a universal language, only one that is as expressive as reasonably possible. %only in a subset of all constructs that can be specified in a mutually orthogonal manner by a semantics of a certain prototypic ``shape''. We will make this more specific later.

\section{Summary of Judgements}\label{sec:verse}
To situate ourselves within a formal framework, let us begin with a brief summary of how Verse is organized and specified. %Verse consists of a \emph{module language} atop a \emph{core language}.

@@

The core language is organized much like the first stage of a type-directed compiler (e.g. the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL). The main judgements in the specification of the EL take the following form  (omitting various contexts for now):\footnote{Note that we use the terms ``expression'' and ``type'' without qualification to refer to  external expressions and types, respectively. }
\\[1ex]
$\begin{array}{ll}
\textbf{Judgement Form} & \textbf{Pronunciation}\\
\vdash e \Rightarrow \sigma \leadsto \iota & \text{Expression $e$ synthesizes type $\sigma$ and has translation $\iota$.}\\
\vdash e \Leftarrow \sigma \leadsto \iota & \text{Expression $e$ analyzes against type $\sigma$ and has translation $\iota$.}\\
\vdash \sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is a type with translation $\tau$.}
\end{array}
$\\[1ex]
Note that the expression typing judgements are \emph{bidirectional}, i.e. we make a judgemental distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}. This is to  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce. Like Scala, we intentionally do not attempt to support global type inference.%, primarily because it is not technically compatible with the primitives we will introduce (but also because we believe that compilers for languages that support only local type inference generate clearer error messages).

@@

Features like state, exceptions, simple concurrency primitives, scalars, arrays and others characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and in some cases this would affect the shape of the internal semantics, but we omit them for simplicity. %Features like record types and labeled sum types, on the other hand, are not built in.

>>>>

%Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives. %New primitives, and thus new dialects of these languages, should arise very rarely. 
%New primitives (and new \emph{dialects} of the language). %Extending a general-pandurpose language with new primitives, i.e.  forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare. %Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality. %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
%Many languages claim to achieve this ideal. 
%For example, 
%If we measure the generality of a language by how frequently  dialects of the language emerge, 
%all major contemporary languages seem to leave room for improvement. 
%General-purpose programming languages are a dime a. 
%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it. Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in. 
Functional programming languages like Standard ML (SML), OCaml and Haskell descend from mathematically elegant typed lambda calculi, diverging mainly in that they consider various pragmatic issues alongside  fundamental metatheoretic issues in their design.  
For example, they build in record types, which generalize the nullary and binary product types more common in minimal calculi, because labeled components can be cognitively useful to humans. Similarly, they build in derived syntax for commonly used constructs like lists.% concision and readability. 
%Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML. to be useful in practice, it must present a decidedly more elaborate concrete syntax and type structure to programmers, designed considering various human factors in addition to fundamental metatheoretic issues. For example, while it is quite common to study minimal calculi that build in only nullary and binary product types, Standard ML (SML) builds in record types, because labels are cognitively useful. %syntactic and semantic primitives that capture common idioms naturally and precisely.

Ideally, a limited number of ``embellishments'' like these would suffice to transition from a minimal calculus to a practical ``full-scale'' programming language. The community around the resulting language would then be able to direct its efforts exclusively toward the development of useful libraries. Unfortunately, a stable language design that achieves this ideal has yet to emerge, as evidenced by the diverse array of {dialects} that continue to proliferate around all major contemporary languages. % For example, let us consider the many dialects of ML. Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with recursive sum and product types, they all introduce various primitive syntactic and semantic extensions and tweaks of their own design (we cover several examples below). 
%Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
Indeed, tools that assist with the construction of  ``domain-specific'' language dialects seem only to be increasingly popular. 
{This calls for an investigation}: why are programmers and researchers still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major functional languages?
%Let us investigate why this is.
%What kinds of constructs are not expressible in languages like SML?

@@

Perhaps the most common reason may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal. To decrease syntactic cost, library providers are often tempted to construct \emph{derived syntactic dialects}, i.e. dialects that are specified by context-free elaboration to the existing language. For example, Ur/Web is a  derived syntactic dialect of Ur (a language that descends from ML) that builds in derived syntax (colloquially, ``syntactic sugar'') for SQL queries, HTML elements and other datatypes used for web programming \cite{conf/popl/Chlipala15}. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
Many other types of data can similarly benefit from the availability of specialized derived syntax. For example, we will consider regular expression patterns expressed using abstract data types (as a mode of use of the ML module system) in Sec. \ref{sec:syntax}. %Though we will be able to express  precisely the semantics that we seek,   the usual approach of using dynamic string parsing to introduce a pattern will leave something to be desired syntactically. 
% An extension of Moscow ML provided syntax for monads encoded using the module system, inspired by the  syntax built in to Haskell.\todo{asked for reference, if not found will say that "Harper reports..."} %In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g. derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}. 
Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} are designed to lower the engineering costs of constructing derived syntactic dialects in situations like these, contributing to their continued proliferation. %(we will consider them further in Sec. \ref{sec:syntax}). % contributing to their proliferation.

@@

More advanced dialects define new static type structure, going beyond what is possible by context-free elaboration to the existing language. Record types themselves are perhaps  the simplest example -- they cannot be expressed by context-free elaboration to a language with only nullary and binary products. Various dialects of ML and Haskell have explored  ``record-like'' primitives that go even further, supporting functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
, methods, prototypic dispatch and other semantic ``embellishments'' that cannot be expressed by context-free elaboration to a language with only standard record types (we will detail one  example in Sec. \ref{sec:metamodules}). The OCaml dialect of ML primitively builds in ``polymorphic variants'', ``open datatypes'' and  logic for typechecking operations that use format strings like $\mathtt{sprintf}$ \cite{ocaml-manual}. ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}. ML5 builds in high-level primitives for distributed programming based on a modal lambda calculus \cite{Murphy:2007:TDP:1793574.1793585}. Manticore \cite{conf/popl/FluetRRSX07} and Alice ML  \cite{AliceLookingGlass} build in parallel programming primitives with a more elaborate type structure than is found in their respective core calculi. 
MLj builds in the type structure of the Java object system (motivated by a desire to interface safely and naturally with Java libraries) \cite{Benton:1999:IWW:317636.317791}. Other dialects do the same for other foreign languages, e.g. Furr and Foster describe a dialect of OCaml that builds in the type structure of C \cite{Furr:2005:CTS:1065010.1065019}. Tools like proof assistants and logical frameworks can be used to specify and reason metatheoretically about dialects like these, and tools like compiler generators lower the cost of implementing them, again contributing to their continued proliferation.

@@

% express record types as syntactic sugar over the simply-typed lambda calculus with  binary product types.\footnote{Pairs can of course be expressed as syntactic sugar atop records, though one could argue that using binary products as the more primitive concept is simpler.} The static semantics need to be extended with new type and term operators. However, the simplest way to express the dynamic semantics of the newly introduced term operators is by translation to nested binary products, so we can leave the operational semantics alone. \todo{fill this out} %For example, there are dozens of constructs that go by the name of ``records'' in various languages, each defined by a slightly different collection of primitive operations. \todo{examples} %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}. Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally. 
% test 
The reason why this proliferation of language dialects should be considered alarming is that it is, in an important sense, anti-modular: a library written in one dialect cannot, in general, safely and idiomatically interface with a library written in another dialect. %At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,  
As MLj demonstrates, addressing the interoperability problem requires somehow ``combining'' the  dialects into a single language. In the most general setting where the dialects in question can be specified by judgements of arbitrary form, this is not a well-defined notion. Even if we restrict our interest to dialects specified using formalisms that do operationalize some notion of dialect combination, there is generally no guarantee that the combined dialect will conserve syntactic and semantic properties that can be established about the dialects in isolation. %In other words, any putative ``combined language'' must formally be considered a  distinct system for which one must derive essentially all metatheorems of interest anew, guided only informally by those derived for the dialects individually. %There is no well-defined mechanism for constructing such a ``combined language'' in general. 
For example, consider two derived syntactic dialects specified using context-free grammars, one specifying derived syntax for finite mappings, the other specifying a similar syntax for ordered finite mappings. Though each can be shown to have an unambiguous concrete syntax in isolation, when their grammars are na\"ively  combined by, for example, Camlp4,  ambiguities  arise.  % would generally attempt to combine their semantics, particularly when they are specified in very different ways. %the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects. 
%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. %A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable. 
%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
Due to this paucity of modular reasoning principles, language dialects are not practical for software development ``in the large''. %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.

Dialect designers must ultimately take a less direct approach to have an impact on large-scale software development -- they must convince the designers in control of comparatively popular languages, like Ocaml and Scala, to include some suitable variant of the primitives they espouse into backwards compatible language revisions. %These decisions are increasingly influenced by community processes, e.g. the Scala Improvement Process.  %This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g. the Scala Improvement Process). 
%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
%Over time, accepting such extensions has caused these languages to balloon in size. 
This \emph{ad hoc} approach is not sustainable, for two main reasons. First, as suggested by the diversity of examples given above, there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}). Second, primitives introduced earlier in a language's lifespan end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms. A related problem is that primitives cannot be removed or changed without breaking backwards compatibility. %rimitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still quite often left languishing in impractical ``toy'' dialects. 
Recalling the words of  Reynolds, which are as relevant today as they were almost half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition. %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.

\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
and generality in language design is a serious unsolved 
problem.}%\begin{flushright}--- John Reynolds (1970)\end{flushright}
\end{quote}

\newpage %A measured approach to incorporating new primitives into a general-purpose programming language is  sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below. 
This leaves the subset of the  language design community interested in keeping general-purpose languages small and free of \emph{ad hoc} primitives with two possible paths forward. One, exemplified (arguably) by SML, is to simply eschew the introduction of specialized syntax and type structure and settle on the existing primitives, which can be said to sit at a ``sweet spot'' in the overall language design space (accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost). % that this implies.  
The other is to search for even more general primitives that reduce these seemingly \emph{ad hoc} primitives to modularly composable library constructs. % so that they can be evaluated on their individual merits by programmers and used together without the possibility of conflict. % This should render the construction of new dialects increasingly unnecessary. 
Encouragingly, primitives of this sort do occasionally arise. For example, a recent revision of Ocaml added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}. Using GADTs, Ocaml was able to move some of the \emph{ad hoc} machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into a library (however, syntactic machinery remains built in).

@@

%\vspace{-10px}
\section{Proposed Contributions}
Our broad aim in the work being proposed is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes more radical steps down the second path just described by giving library providers the ability to directly express derived syntax and type structure of a variety of designs in a safe and modularly composable manner. %This eliminates the needs to build in fewer \emph{ad hoc} constructs and dialects are less frequently necessary. %thereby reducing the need for language dialects and revisions.

Verse features a module system taken directly from  SML. Unlike SML, the Verse core language is split into a \emph{typed external language} (EL) specified by {type-directed translation} to a minimal \emph{typed internal language} (IL). We formally introduce the semantics of the Verse core language in Sec. \ref{sec:verse}. Our novel contributions will be two primitive constructs that eliminate the need for many others:
\begin{itemize}
\item \textbf{Typed syntax macros} (TSMs), introduced in Sec. \ref{sec:syntax}, eliminate the need to primitively build in derived concrete syntax specific to library constructs, e.g. list syntax as in SML or XML syntax as in Scala and Ur/Web, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax. %We show how TSMs can be invoked explicitly by clients, or, implicitly based on local type inference.
\item \textbf{Metamodules}, introduced in Sec. \ref{sec:metamodules}, eliminate the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other more esoteric constructs that we will consider by giving library providers static hooks directly into Verse's type-directed translation semantics. For example, a library provider can implement the type structure of records by defining a metamodule that:
\begin{enumerate}
\item Introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types and defines its semantics by translation to unary and binary products (which are built in to the internal language).
\item Introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly defines the logic governing their typechecking and translation. 
\end{enumerate}
We will see direct analogies between modules and metamodules in Sec. \ref{sec:metamodules}.
\end{itemize} 
Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because their static semantics will involve the evaluation of user-defined functions that manipulate static representations of types and terms.% static code generation. %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.

The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to decentralize control over aspects of the language's concrete syntax and type structure that, in other contemporary languages, are under the exclusive control of the language designer. If we are not careful, many of the problems discussed earlier that arise when combining language dialects could simply shift into the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer. The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended. We will avoid this somewhat confusing terminology.} Our main technical contributions will come in showing how to address these problems in a principled manner. In particular, syntactic conflicts will be impossible by construction and the semantics will validate code statically generated by TSMs and metamodules to maintain a strong \emph{hygienic type discipline} and, most uniquely, powerful \emph{modular reasoning principles}. In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and clients will be able to use them safely in any program context and in any combination, without the possibility of conflict.\footnote{This is not quite true -- simple naming conflicts can arise. We will tacitly assume that they are being avoided extrinsically, e.g. by using a URI-based naming scheme as in the Java ecosystem.}

@@

Finally, we will not be interested here in languages that feature full-spectrum dependent types, which blur the phase separation between compile-time and run-time, though we conjecture that the primitives we introduce could be introduced into languages like Gallina (the ``external language'' of the Coq proof assistant) with some reasonable modifications. Verse is designed for use in situations where ML, Haskell or Scala would be suitable today. %Our interest is in making sure that the standard library is no more privileged than any other library. % Our interest is not in creating a universal language, only one that is as expressive as reasonably possible. %only in a subset of all constructs that can be specified in a mutually orthogonal manner by a semantics of a certain prototypic ``shape''. We will make this more specific later.

\section{Overview of Judgements}\label{sec:verse}
To situate ourselves within a formal framework, let us begin with a brief overview of how Verse is organized and specified. %Verse consists of a \emph{module language} atop a \emph{core language}.

@@

The Verse core language will be the focus of our formal efforts. It is organized much like the first stage of a type-directed compiler (e.g. the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL). The main judgements in the specification of the EL take the following form  (omitting various contexts for now):\footnote{Note that we will use the terms ``expression'' and ``type'' without qualification to refer to  external expressions and types, respectively. }
\\[1ex]
$\begin{array}{ll}
\textbf{Judgement Form} & \textbf{Pronunciation}\\
\vdash e \Rightarrow \sigma \leadsto \iota & \text{Expression $e$ synthesizes type $\sigma$ and has translation $\iota$.}\\
\vdash e \Leftarrow \sigma \leadsto \iota & \text{Expression $e$ analyzes against type $\sigma$ and has translation $\iota$.}\\
\vdash \sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is a type with translation $\tau$.}
\end{array}
$\\[1ex]
Note that the expression typing judgements are \emph{bidirectional}, i.e. we make a judgemental distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}. This is to  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we aim to introduce. Like Scala, we intentionally do not attempt to support global type inference.%, primarily because it is not technically compatible with the primitives we will introduce (but also because we believe that compilers for languages that support only local type inference generate clearer error messages).

@@

Features like state, exceptions, simple concurrency primitives, scalars, arrays and others characteristic of a first-stage compiler intermediate language would also be included in the IL in practice, and in some cases this would affect the shape of the internal semantics, but we will omit them for simplicity. %Features like record types and labeled sum types, on the other hand, are not built in.
